Natural selection in the Health and Retirement Study

### 2023-12-04 hrs-selection.qmd
Cleanups

@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from \@becker1976child. The theory has two components.

@@

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey \[\@hrsrandfam2023; \@hrsrand2023\]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The HRS also provides precalculated polygenic scores \[\@hrspgs2020\] for both black and (non-hispanic) white participants, so we can check whether the theory predicts patterns of natural selection in both these ethnicities.

@@

The HRS contains weights which match survey respondents to the US population. We use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, we weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012. Statistical tests are adjusted for clustering and stratification using the R "survey" package \[\@lumleysurvey2023\].

@@

We estimated coefficients of PGS on RLRS among black and white respondents separately. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix \@fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`. The scores are age at first birth, educational attainment and attention deficit disorder. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Survey bootstraps \[\@canty1999resampling\] are used so as to make inferences from the sample of respondents.

@@

|               | Low group             | High group            | Difference            |
|-----------------|-------------------|-------------------|-------------------|
| Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
| Income        | `r tbl$income_med[1]` | `r tbl$income_med[2]` | `r tbl$income_med[3]` |
| Marriage      | `r tbl$married[1]`    | `r tbl$married[2]`    | `r tbl$married[3]`    |
| Age 1st birth | `r tbl$agefb[1]`      | `r tbl$agefb[2]`      | `r tbl$agefb[3]`      |
| Birth year    | `r tbl$born[1]`       | `r tbl$born[2]`       | `r tbl$born[3]`       |

@@

where \$h\^2\$ is the heritability of the PGS target phenotype and \$R\^2\_{PGS}\$ is the \$R\^2\$ of the measured PGS on the target phenotype \[\@becker2021resource\]. Moving from error correction in a univariate model to a multivariate model requires a slightly more complex correction, although it uses the same parameters. To do this we use the error-corrected estimator developed by \@becker2021resource.

@@

system("rm pgi_correct_data.txt")

>>>>

@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components.

@@

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so we can check whether the theory predicts patterns of natural selection in both these ethnicities.

@@

The HRS contains weights which match survey respondents to the US population. We use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, we weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012. Statistical tests are adjusted for clustering and stratification using the R "survey" package [@lumleysurvey2023].

@@

We estimated coefficients of PGS on RLRS among black and white respondents separately. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`. The scores are age at first birth, educational attainment and attention deficit disorder. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.

@@

|               | Low group             | High group            | Difference            |
|------------------|------------------|------------------|------------------|
| Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
| Income        | `r tbl$income_med[1]` | `r tbl$income_med[2]` | `r tbl$income_med[3]` |
| Marriage      | `r tbl$married[1]`    | `r tbl$married[2]`    | `r tbl$married[3]`    |
| Age 1st birth | `r tbl$agefb[1]`      | `r tbl$agefb[2]`      | `r tbl$agefb[3]`      |
| Birth year    | `r tbl$born[1]`       | `r tbl$born[2]`       | `r tbl$born[3]`       |

@@

where \$h\^2\$ is the heritability of the PGS target phenotype and \$R\^2\_{PGS}\$ is the \$R\^2\$ of the measured PGS on the target phenotype [@becker2021resource]. Moving from error correction in a univariate model to a multivariate model requires a slightly more complex correction, although it uses the same parameters. To do this we use the error-corrected estimator developed by \@becker2021resource.

@@

file.remove("pgi_correct_data.txt")
